---
title: "MCMC Convergence Diagnostics"
author: "AnaCoDa Development Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{MCMC Convergence Diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: anacoda.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Overview

After every MCMC run it is essential to verify that the chain has converged
to the posterior distribution before drawing any biological conclusions.
AnaCoDa provides three complementary diagnostic tools:

| Function | Object | What it shows |
|---|---|---|
| `plot(mcmc, ...)` | `Rcpp_MCMCAlgorithm` | Log-posterior / log-likelihood trace over samples |
| `acfMCMC(mcmc)` | `Rcpp_MCMCAlgorithm` | ACF of the log-posterior and log-likelihood traces |
| `acfCSP(parameter)` | `Rcpp_ROCParameter` (etc.) | Per-amino-acid ACF grid for codon-specific parameters |

All three should be inspected together: trace plots reveal trends and slow drift,
while ACF plots reveal the *mixing rate* -- how many MCMC steps separate
effectively independent draws.

## Setup

The examples below assume you have already run MCMC and have the following
objects in your workspace.  See the *AnaCoDa* main vignette for a full
end-to-end workflow.

```{r run-mcmc}
library(AnaCoDa)

genome    <- initializeGenomeObject(file = "genome.fasta",
                                    observed.expression.file = "phi.csv")
parameter <- initializeParameterObject(genome, sphi_init = 1,
                                       num.mixtures = 1,
                                       gene.assignment = rep(1, length(genome)))
model     <- initializeModelObject(parameter, "ROC", with.phi = TRUE)
mcmc      <- initializeMCMCObject(samples = 10000, thinning = 10,
                                  adaptive.width = 50)
runMCMC(mcmc, genome, model, ncores = 4)
```

## Trace plots

### Log-posterior / log-likelihood trace

The simplest first check: does the chain appear to have reached a stationary
distribution?  Upward or downward trends in the last portion of the run suggest
the chain has not yet converged.

```{r trace-mcmc}
# Log-posterior trace with a zoomed inset of the last 10 %
plot(mcmc, what = "LogPosterior", zoom.window = c(9000, 10000))

# Switch to log-likelihood
plot(mcmc, what = "LogLikelihood", zoom.window = c(9000, 10000))
```

### Codon-specific parameter traces

The parameter trace object stores the full sample history for every
codon-specific parameter.  Panels are organised by amino acid.

```{r trace-csp}
trace <- parameter$getTraceObject()

# Mutation parameter (Delta M) for mixture 1
plot(trace, what = "Mutation", mixture = 1)

# Selection parameter (Delta eta) for mixture 1
plot(trace, what = "Selection", mixture = 1)
```

## ACF diagnostics

### Rationale

A well-mixing chain produces autocorrelation values that drop quickly --
ideally within the first five to ten lags -- to inside the dashed
$\pm\,1.96/\sqrt{n}$ confidence band.
Persistent positive autocorrelation beyond lag 20-30 is a warning sign;
values that are still large at `lag.max` indicate the chain is effectively
exploring only a tiny fraction of the posterior per step.

### `acfMCMC`: log-posterior and log-likelihood ACF

`acfMCMC` overlays the ACF of both traces on a single panel, color-coded via
the mixture color scheme.  The `what` argument (default: both) mirrors
`plot.Rcpp_MCMCAlgorithm`.

```{r acfMCMC-both}
# Both traces on one panel (default)
acfMCMC(mcmc, lag.max = 40)
```

```{r acfMCMC-single}
# Log-posterior only
acfMCMC(mcmc, what = "LogPosterior", lag.max = 40)
```

The `samples` argument controls how many samples from the *end* of the trace
are used (default: `10 * log10(N)`).  Use a longer suffix to get a stable ACF
estimate; use a shorter one to focus on the post-warmup behaviour.

```{r acfMCMC-samples}
# Use the last 2000 samples
acfMCMC(mcmc, samples = 2000, lag.max = 40)
```

`acfMCMC` invisibly returns a named list of `acf` objects so the raw values
are available for further analysis:

```{r acfMCMC-return}
acf.out <- acfMCMC(mcmc, plot = FALSE)
# Lag-1 autocorrelation of the log-posterior
acf.out$LogPosterior$acf[2]
```

### `acfCSP`: codon-specific parameter ACF

`acfCSP` produces a compact multi-panel grid -- one panel per amino acid, one
colored line per synonymous codon -- analogous to the trace layout produced by
`plot.Rcpp_Trace`.

#### Fixed y-scale (default)

Under `scales = "fixed"` all panels share a common y-range and panels in the
same column are merged vertically (zero gap), making it easy to compare
mixing across amino acids at a glance.
Axis labels appear only on the four outer frame edges; interior borders carry
tick marks only.

```{r acfCSP-fixed}
# Mutation parameter, mixture 1, fixed y-scale
acfCSP(parameter, what = "Mutation", mixture = 1, scales = "fixed")
```

```{r acfCSP-selection}
# Selection parameter
acfCSP(parameter, what = "Selection", mixture = 1, scales = "fixed")
```

#### Free y-scale

Use `scales = "free_y"` when you need to inspect the shape of individual
amino-acid ACF curves without the global range compressing small-amplitude
signals.

```{r acfCSP-free}
acfCSP(parameter, what = "Mutation", mixture = 1, scales = "free_y")
```

#### ESS/1k mixing metric

Each legend entry includes an **ESS/1k** value: the effective number of
independent draws per 1000 MCMC steps,

$$\text{ESS/1k} = \frac{1000}{1 + 2\sum_{k \geq 1} \rho_k}$$

where the sum runs over all positive-lag autocorrelations up to the first
non-positive lag.  This metric is *independent of how many samples you passed*
to `acfCSP`, making it comparable across runs with different chain lengths or
thinning intervals.

As a rough guide:

| ESS/1k | Interpretation |
|---|---|
| > 500 | Good mixing; chain explores efficiently |
| 100 -- 500 | Moderate autocorrelation; may need more samples |
| 10 -- 100 | High autocorrelation; consider wider adaptive proposals |
| < 10 | Very poor mixing; chain is nearly stuck |

#### Return value

`acfCSP` invisibly returns a nested list (amino acid → codon → `acf` object)
for downstream use:

```{r acfCSP-return}
acf.out <- acfCSP(parameter, what = "Mutation", mixture = 1, plot = FALSE)

# Lag-1 autocorrelation for each GCN codon (Alanine)
lapply(acf.out[["A"]], function(a) as.numeric(a$acf)[2])
```

## Worked interpretation example

A healthy 10,000-sample ROC run with default adaptive width typically shows:

- Log-posterior ACF dropping below the CI band by lag 5--15.
- Mutation ($\Delta M$) ACF: most codons clear the band by lag 10--20;
  rare codons (1--2 observations in the genome) may remain elevated.
- Selection ($\Delta\eta$) ACF: similar pattern to mutation but potentially
  slower for codons under weak selection.

If many codons show ESS/1k $< 50$ after a 10,000-step run, consider:

1. **Increasing `adaptive.width`** in `initializeMCMCObject` to allow wider
   adaptive proposals.
2. **Running longer** and discarding a larger burn-in fraction.
3. **Inspecting the trace plots** to check for multi-modal behaviour or
   sudden jumps that would inflate apparent autocorrelation.

## References
