---
title: "Filtering RFP and Phi Data"
author: "Michael Gilchrist"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Filtering RFP and Phi Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

## Purpose

`filterRFPData()` filters a tidy long-format ribosome footprint (RFP)
data frame and its matching wide-format phi (expression) data frame
by gene length, read counts, pause-site outliers, position trimming,
gene-ID exclusion, and random gene subsampling. It is intended as a
pre-processing step before loading data into AnaCoDa for fitting PA,
PANSE, FONSE, or any other RFP-based model.

The function returns a list `list(rfp, phi)` with both tables filtered
to a consistent surviving gene set.

## Input formats

- **RFP table:** tidy long format with columns `GeneID`, `Position`,
  `Codon`, `RFPCount`. One row per (gene, codon position).
- **Phi table:** wide format with a `GeneID` column plus one or more
  expression value columns.

## Filter order

Filters are applied in a fixed order regardless of argument order:

1. Exclude by gene ID (`exclude_genes`)
2. Minimum gene length (`min_length_gene`)
3. Minimum total read count (`min_read_counts`), global or per-half
4. Minimum read count per position (`min_read_counts_per_pos`),
   global or per-half
5. Pause-site removal (`remove_pause_sites`)
6. Position ceiling (`last_position_to_include`)
7. 5' trim (`trim_5`), with renumbering
8. 3' trim (`trim_3`)
9. Random subsample (`num_genes_to_include`, optionally seeded)

After RFP filtering completes, the phi table is subset and reordered
to match `unique(rfp$GeneID)`.

## Example

```{r example, eval = FALSE}
library(AnaCoDa)

rfp <- read.csv(system.file("extdata", "test_rfp.csv",
                            package = "AnaCoDa"),
                stringsAsFactors = FALSE)
phi <- read.csv(system.file("extdata", "test_phi.csv",
                            package = "AnaCoDa"),
                stringsAsFactors = FALSE)

out <- filterRFPData(
  rfp, phi,
  min_length_gene    = 5,     # drop short genes
  min_read_counts    = 100,   # drop low-coverage genes
  remove_pause_sites = 2,     # drop genes with Z-score > 2 outlier
  trim_5             = 5,     # drop the first 5 codons
  trim_3             = 5      # drop the last 5 codons
)

nrow(out$rfp)
nrow(out$phi)
```

## CLI wrapper

A command-line wrapper that mirrors the function's interface and
handles I/O is shipped with the package at:

```{r cli-path, eval = FALSE}
system.file("scripts", "createSubSampled.R", package = "AnaCoDa")
```

It requires the `argparse` and `readr` packages. Run with `Rscript`:

```sh
Rscript createSubSampled.R \
  --input rfp.csv --phi phi.csv \
  --output_dir out/ \
  --output_rfp filtered_rfp.csv \
  --output_phi filtered_phi.csv \
  --min_length_gene 100 \
  --min_read_counts 200 \
  --remove_pause_sites 3
```

The wrapper also accepts legacy column names `gene` (in place of
`GeneID` in the RFP table) and `ORF` (in place of `GeneID` in the
phi table), renaming them before calling the library function.

## Notes

- `min_read_counts` and `min_read_counts_per_pos` are mutually
  exclusive; specifying both is an error.
- Combining `last_position_to_include` and `trim_5` is supported but
  emits a warning, since `last_position_to_include` is applied first
  and `trim_5` then drops additional codons from what remains.
- `apply_filter_to_halves = TRUE` evaluates the active read-count
  threshold independently on the 5' and 3' halves; both halves must
  pass for the gene to be kept.
- `seed` sets the global RNG state via `set.seed()` before random
  subsampling, making the subsample reproducible.
