---
title: "AnaCoDa: FONSE Model for Nonsense Error Estimation"
author: "Cedric Landerer, Alexander L. Cope & Michael A. Gilchrist"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: anacoda.bib
vignette: >
  %\VignetteIndexEntry{AnaCoDa: FONSE Model for Nonsense Error Estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

> **Status**: The FONSE model is unpublished and currently in beta testing.
> Results should be interpreted with caution and validated against simulated data
> before use in publications.

AnaCoDa's **FONSE** (First Order approximation of NonSense Error) model estimates
codon-specific nonsense error probabilities from coding sequences.
Nonsense errors occur when a translating ribosome prematurely terminates at a
sense codon, producing a truncated, non-functional protein.
The theory predicts that selection against nonsense errors is strongest toward
the 3'-end of transcripts (errors near the end waste less energy), which produces
the commonly observed enrichment of rare, inefficient codons toward the 5'-end.

FONSE models this with a position-dependent cost function
$$\beta(k) = a_1 + a_2 \cdot k$$
where $k$ is the codon position and $a_1$ is the initiation cost hyperparameter
(estimated from data). The constant $a_2 = 4.0$ is currently fixed.

FONSE estimates three sets of parameters:

| Parameter | Symbol | Description |
|-----------|--------|-------------|
| Mutation bias | $\Delta M$ | Log-odds of codon usage relative to reference codon, absent selection |
| Nonsense error coefficient | $\Delta\Omega$ | Selection against nonsense error at each codon |
| Initiation cost | $a_1$ | Indirect cost of translation initiation (hyperparameter) |
| Synthesis rate | $\phi$ | Per-gene expression level |

For each non-reference codon $i$ of amino acid $A$, the codon probability is
proportional to $\exp(-\Delta M_i - \Delta\Omega_i \cdot \phi \cdot \beta(k))$,
with the reference codon (last alphabetically per amino acid) fixed at 1.

## Input data

FONSE uses the same input as ROC: coding sequences in FASTA format.

```{r, eval = FALSE}
library(AnaCoDa)
genome_file <- system.file("extdata", "genome.fasta", package = "AnaCoDa")
genome <- initializeGenomeObject(file = genome_file)
```

For observed expression data (RNA-seq, mass spectrometry), supply a CSV file
with columns `ORF,<data_col_1>[,<data_col_2>,...]`:

```{r, eval = FALSE}
expression_file <- "expression.csv"
genome <- initializeGenomeObject(file = genome_file,
                                 observed.expression.file = expression_file,
                                 match.expression.by.id = TRUE)
```

## Initializing the parameter object

Specify `model = "FONSE"` when initializing the parameter object.
The `init.initiation.cost` argument sets the starting value for $a_1$
(default is 4).

```{r, eval = FALSE}
n_genes      <- length(genome)
sphi_init    <- 1
num_mixtures <- 1
gene_assignment <- rep(1, n_genes)

parameter <- initializeParameterObject(
  genome             = genome,
  sphi               = sphi_init,
  num.mixtures       = num_mixtures,
  gene.assignment    = gene_assignment,
  model              = "FONSE",
  init.initiation.cost = 4
)
```

### Using multiple mixtures

FONSE supports the same mixture definitions as ROC: `"allUnique"` (default),
`"mutationShared"`, and `"selectionShared"`.

```{r, eval = FALSE}
sphi_init    <- c(1, 1)
num_mixtures <- 2
gene_assignment <- sample(c(1, 2), size = n_genes, replace = TRUE,
                           prob = c(0.5, 0.5))

parameter <- initializeParameterObject(
  genome             = genome,
  sphi               = sphi_init,
  num.mixtures       = num_mixtures,
  gene.assignment    = gene_assignment,
  model              = "FONSE",
  mixture.definition = "allUnique",
  init.initiation.cost = 4
)
```

### Initializing from files

If you have prior estimates of $\Delta M$ or $\Delta\Omega$ (e.g., from a
previous run), load them with `initMutationCategories` and
`initSelectionCategories`:

```{r, eval = FALSE}
parameter$initMutationCategories(c("mutation_cat1.csv"), numCategories = 1,
                                  fix = FALSE)
parameter$initSelectionCategories(c("selection_cat1.csv"), numCategories = 1,
                                   fix = FALSE)
```

## Initializing the model and MCMC objects

```{r, eval = FALSE}
model <- initializeModelObject(parameter, "FONSE")

mcmc <- initializeMCMCObject(
  samples        = 5000,
  thinning       = 10,
  adaptive.width = 50,
  est.expression = TRUE,
  est.csp        = TRUE,
  est.hyper      = TRUE
)
```

## Running the MCMC

```{r, eval = FALSE}
runMCMC(mcmc = mcmc, genome = genome, model = model, ncores = 4)
```

## Diagnostic plots

After the MCMC completes, inspect the traces to assess convergence.

### Log-posterior trace

```{r, eval = FALSE}
plot(mcmc)
```

### Initiation cost ($a_1$) trace

```{r, eval = FALSE}
trace <- parameter$getTraceObject()
plot(trace, what = "InitiationCost")
```

### Codon-specific parameter traces

```{r, eval = FALSE}
# Mutation bias traces for all amino acids
plotCodonSpecificParameters(parameter, what = "Mutation")

# Nonsense error coefficient traces
plotCodonSpecificParameters(parameter, what = "Selection")
```

## Extracting posterior estimates

### Initiation cost ($a_1$)

```{r, eval = FALSE}
trace    <- parameter$getTraceObject()
a1_trace <- trace$getInitiationCostTrace()

# Discard burn-in (first 50% of samples)
n_samples <- length(a1_trace)
a1_post   <- a1_trace[ceiling(n_samples / 2):n_samples]

cat("a1 posterior mean:", mean(a1_post), "\n")
cat("a1 95% CI:        ", quantile(a1_post, c(0.025, 0.975)), "\n")
```

### Codon-specific parameters ($\Delta M$ and $\Delta\Omega$)

`getCSPEstimates` returns a list with posterior means and credible intervals
for both parameter types.

```{r, eval = FALSE}
csp <- getCSPEstimates(parameter, samples = 1000, mixture = 1)

# Mutation bias (DeltaM)
head(csp$Mutation)

# Nonsense error coefficients (DeltaOmega)
head(csp$Selection)
```

To obtain estimates for a single codon:

```{r, eval = FALSE}
# Posterior mean of DeltaOmega for codon "AAA" in mixture 1
# paramType = 1 for selection (DeltaOmega)
omega_aaa <- parameter$getCodonSpecificPosteriorMean(
  mixtureElement    = 1,
  samples           = 1000,
  codon             = "AAA",
  paramType         = 1,
  withoutReference  = TRUE
)
```

### Synthesis rates ($\phi$)

```{r, eval = FALSE}
phi_estimates <- getExpressionEstimates(parameter, genome, samples = 1000)
head(phi_estimates)
```

## Fixing parameters

To test hypotheses or speed up sampling, individual parameter sets can be held
fixed during the MCMC run.

```{r, eval = FALSE}
# Fix mutation bias at current values
parameter$fixDM()

# Fix nonsense error coefficients at current values
parameter$fixDOmega()

# Fix the initiation cost at its current value
parameter$fixedInitiationCost()
```

Alternatively, pass `fix = TRUE` when loading categories from files to fix
them immediately:

```{r, eval = FALSE}
parameter$initMutationCategories(c("mutation_cat1.csv"), numCategories = 1,
                                  fix = TRUE)
```

## Restart files

Use restart files to resume an interrupted run or extend a completed one.

```{r, eval = FALSE}
# Enable restart file output every 100 MCMC samples
setRestartSettings(mcmc, filename = "fonse_restart", samples = 100,
                    overwrite = FALSE)

# Resume from a restart file
parameter_resumed <- initializeParameterObject(
  init.with.restart.file = "fonse_restart.rst"
)
model_resumed <- initializeModelObject(parameter_resumed, "FONSE")
mcmc_resumed  <- initializeMCMCObject(samples = 5000, thinning = 10,
                                       adaptive.width = 50,
                                       est.expression = TRUE,
                                       est.csp        = TRUE,
                                       est.hyper      = TRUE)
runMCMC(mcmc = mcmc_resumed, genome = genome, model = model_resumed, ncores = 4)
```

## Saving and loading MCMC objects

```{r, eval = FALSE}
writeMCMCObject(mcmc, file = "fonse_mcmc.Rda")
mcmc_loaded <- loadMCMCObject(file = "fonse_mcmc.Rda")
```

## Known limitations

- **Unpublished model**: FONSE has not undergone peer review.  Results should
  be cross-validated against simulations before use in publications.
- **Fixed $a_2$**: The positional scaling parameter $a_2 = 4.0$ is hard-coded
  throughout the model.  It is not estimated from data.
- **Computational cost**: Because FONSE iterates over codon *positions* rather
  than counts, it is more expensive per gene than ROC for long sequences.
