---
title: "AnaCoDa: PA and PANSE Models for Ribosome Footprinting Data"
author: "Cedric Landerer, Alexander L. Cope & Michael A. Gilchrist"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: anacoda.bib
vignette: >
  %\VignetteIndexEntry{AnaCoDa: PA and PANSE Models for Ribosome Footprinting Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

AnaCoDa provides two models for analyzing ribosome footprinting (RFP) data: the
**PA** (PAusing time) model and its extension **PANSE** (PAusing time plus NonSense
Error).
Both models use a negative binomial likelihood to describe per-codon RFP counts,
estimating pausing shape ($\alpha$), elongation rate ($\lambda'$), and gene-level
expression ($\psi$) for each codon and gene in the genome.

PANSE additionally estimates per-codon nonsense error (NSE) rates, which reflect the
probability that a translating ribosome terminates prematurely at a given codon.
These NSE rates modulate the effective expression seen at downstream positions.

## Input data format

PA and PANSE models require ribosome footprinting count data rather than coding
sequences.
The input file is a comma-separated table with one row per gene-codon combination:

```
GeneID,count_1,count_2,Codon
YAL001C,110,73,AAA
YAL001C,28,22,AAC
...
```

Each row provides the raw and normalized RFP counts (or two independent count
measurements) for a single codon in a single gene, along with the codon identity.
Stop codons are excluded.
The `fasta = FALSE` argument to `initializeGenomeObject` signals that the file
contains RFP data rather than coding sequences.

A 500-gene example dataset is included with the package:

```{r, echo = TRUE, eval = FALSE}
rfp_file <- system.file("extdata", "pa_rfpdata.csv", package = "AnaCoDa")
```

## Running the PA model

### Step 1: Load genome

```{r, echo = TRUE, eval = FALSE}
library(AnaCoDa)

rfp_file <- system.file("extdata", "pa_rfpdata.csv", package = "AnaCoDa")
genome <- initializeGenomeObject(file = rfp_file, fasta = FALSE)
```

### Step 2: Initialize parameter object

The PA parameter object requires specifying an initial value for the hyperparameter
$s_\phi$ (`sphi`), which controls the lognormal prior on gene-level expression
$\psi$.
With a single gene-set (no mixture) all genes share the same codon-specific
parameters $\alpha$ and $\lambda'$.

```{r, echo = TRUE, eval = FALSE}
sphi_init    <- 2
numMixtures  <- 1
geneAssignment <- rep(1, length(genome))

parameter <- initializeParameterObject(
  genome          = genome,
  sphi            = sphi_init,
  num.mixtures    = numMixtures,
  gene.assignment = geneAssignment,
  model           = "PA",
  split.serine    = TRUE,
  mixture.definition = "allUnique"
)
```

### Step 3: Initialize model and MCMC objects

```{r, echo = TRUE, eval = FALSE}
model <- initializeModelObject(parameter, "PA")

mcmc <- initializeMCMCObject(
  samples        = 1000,
  thinning       = 10,
  adaptive.width = 50,
  est.expression = TRUE,
  est.csp        = TRUE,
  est.hyper      = TRUE
)
```

### Step 4: Run MCMC

`runMCMC` modifies `mcmc` and `parameter` in place; no return value is needed.

```{r, echo = TRUE, eval = FALSE}
runMCMC(mcmc, genome, model)
```

## Saving and reloading results

Parameter and MCMC objects can be written to disk and reloaded in a later session:

```{r, echo = TRUE, eval = FALSE}
writeParameterObject(parameter, file = "pa_parameter.Rdat")
writeMCMCObject(mcmc, file = "pa_mcmc.Rdat")

parameter <- loadParameterObject(file = "pa_parameter.Rdat")
mcmc      <- loadMCMCObject(file = "pa_mcmc.Rdat")
```

## Diagnostic plots

### Log-posterior trace

Inspect convergence by plotting the log-posterior over samples.
Discarding an initial burn-in (here 30% of samples) removes the transient
at the start of the chain.

```{r, echo = TRUE, eval = FALSE}
logPost <- mcmc$getLogPosteriorTrace()
burnin  <- ceiling(length(logPost) * 0.3)
logPost_post_burnin <- logPost[burnin:length(logPost)]

plot(logPost_post_burnin, type = "l",
     main = paste("log(Posterior), mean =", round(mean(logPost_post_burnin), 1)),
     xlab = "Sample", ylab = "log(Posterior)")
grid(NULL, NULL, lty = 6, col = "cornsilk2")
```

### Hyperparameter traces

```{r, echo = TRUE, eval = FALSE}
trace <- parameter$getTraceObject()
plot(trace, what = "Sphi",             main = "s_phi hyperparameter")
plot(trace, what = "MixtureProbability", main = "Mixture probability")
plot(trace, what = "ExpectedPhi",      main = "Expected expression (psi)")
```

### Codon-specific parameter traces

Trace plots for $\alpha$ and $\lambda'$ across all codons help assess per-codon
convergence.

```{r, echo = TRUE, eval = FALSE}
plot(trace, what = "Alpha",       mixture = 1)
plot(trace, what = "LambdaPrime", mixture = 1)
```

### Posterior means and credible intervals

After discarding burn-in, extract posterior means and 95% credible intervals for
each codon's $\alpha$ and $\lambda'$:

```{r, echo = TRUE, eval = FALSE}
samples    <- 1000
codonList  <- codons()          # 61 sense codons
cat        <- 1                 # mixture category index
burnin_idx <- samples * 0.5    # use posterior half

alpha_mean <- numeric(61)
alpha_ci   <- matrix(0, nrow = 61, ncol = 2)
lp_mean    <- numeric(61)
lp_ci      <- matrix(0, nrow = 61, ncol = 2)

for (i in seq_along(codonList)) {
  codon <- codonList[i]

  alpha_mean[i] <- parameter$getCodonSpecificPosteriorMean(cat, burnin_idx, codon, 0, FALSE)
  alpha_trace   <- trace$getCodonSpecificParameterTraceByMixtureElementForCodon(1, codon, 0, FALSE)
  alpha_ci[i, ] <- quantile(alpha_trace[burnin_idx:samples], probs = c(0.025, 0.975))

  lp_mean[i]  <- parameter$getCodonSpecificPosteriorMean(cat, burnin_idx, codon, 1, FALSE)
  lp_trace    <- trace$getCodonSpecificParameterTraceByMixtureElementForCodon(1, codon, 1, FALSE)
  lp_ci[i, ]  <- quantile(lp_trace[burnin_idx:samples], probs = c(0.025, 0.975))
}

# Plot alpha credible intervals
plot(NULL, NULL,
     xlim = c(1, 61), ylim = range(alpha_ci),
     main = "Posterior mean and 95% CI: Alpha",
     xlab = "Codon", ylab = "Alpha", axes = FALSE)
confidenceInterval.plot(x = 1:61, y = alpha_mean, sd.y = alpha_ci)
axis(2)
axis(1, at = 1:61, labels = codonList, tck = 0.02, las = 2, cex.axis = 0.6)

# Plot lambda-prime credible intervals
plot(NULL, NULL,
     xlim = c(1, 61), ylim = range(lp_ci),
     main = "Posterior mean and 95% CI: LambdaPrime",
     xlab = "Codon", ylab = "LambdaPrime", axes = FALSE)
confidenceInterval.plot(x = 1:61, y = lp_mean, sd.y = lp_ci)
axis(2)
axis(1, at = 1:61, labels = codonList, tck = 0.02, las = 2, cex.axis = 0.6)
```

Pausing times (in arbitrary units) are proportional to $\alpha \cdot \lambda'$:

```{r, echo = TRUE, eval = FALSE}
pausing_times <- alpha_mean * lp_mean
barplot(pausing_times, names.arg = codonList,
        main = "Estimated pausing times per codon",
        las = 2, cex.names = 0.6)
```

## PANSE model

> **Status**: The PANSE model is unpublished and currently in beta testing.
> Results should be interpreted with caution and validated against simulated
> data before use in publications.

PANSE extends PA by jointly estimating per-codon nonsense error (NSE) rates.
Ribosomes that drop off at upstream codons reduce the observed RFP counts at
downstream positions; PANSE accounts for this modulation via a cumulative
survival probability.

### Step 1: Load genome (positional mode)

PANSE requires per-position RFP counts; set `positional = TRUE`:

```{r, echo = TRUE, eval = FALSE}
rfp_file   <- system.file("extdata", "pa_rfpdata.csv", package = "AnaCoDa")
genome_nse <- initializeGenomeObject(file = rfp_file, fasta = FALSE,
                                      positional = TRUE)
```

### Step 2: Initialize PANSE parameter object

```{r, echo = TRUE, eval = FALSE}
parameter_nse <- initializeParameterObject(
  genome          = genome_nse,
  sphi            = 2,
  num.mixtures    = 1,
  gene.assignment = rep(1, length(genome_nse)),
  model           = "PANSE",
  split.serine    = TRUE,
  mixture.definition = "allUnique"
)
```

PANSE-specific mixture definitions:

- `"allUnique"` (default): each mixture has its own alpha, lambda, and NSE rates
- `"elongationShared"`: all mixtures share alpha and lambda but have separate NSE rates
- `"nseShared"`: all mixtures share NSE rates but have separate alpha and lambda

### Step 3-4: Model, MCMC, and run

```{r, echo = TRUE, eval = FALSE}
model_nse <- initializeModelObject(parameter_nse, "PANSE")

mcmc_nse <- initializeMCMCObject(
  samples        = 1000,
  thinning       = 10,
  adaptive.width = 50,
  est.expression = TRUE,
  est.csp        = TRUE,
  est.hyper      = TRUE
)

runMCMC(mcmc_nse, genome_nse, model_nse)
```

### NSE rate traces and posterior summaries

NSE rate traces are accessed via `what = "NSERate"` in `plot(trace, ...)`:

```{r, echo = TRUE, eval = FALSE}
trace_nse <- parameter_nse$getTraceObject()
plot(trace_nse, what = "NSERate", mixture = 1)
```

Extract posterior means and credible intervals for NSE rates:

```{r, echo = TRUE, eval = FALSE}
codonList  <- codons()     # 61 sense codons
burnin_idx <- 500          # discard first 50% of 1000 samples

nse_mean <- numeric(61)
nse_ci   <- matrix(0, nrow = 61, ncol = 2)

for (i in seq_along(codonList)) {
  codon <- codonList[i]
  nse_trace   <- trace_nse$getCodonSpecificParameterTraceByMixtureElementForCodon(
                   1, codon, 2, FALSE)
  post_nse    <- nse_trace[burnin_idx:length(nse_trace)]
  nse_mean[i] <- mean(post_nse)
  nse_ci[i, ] <- quantile(post_nse, probs = c(0.025, 0.975))
}

barplot(nse_mean, names.arg = codonList,
        main = "Posterior mean NSE rate per codon",
        las = 2, cex.names = 0.6)
```

`getCSPEstimates` with `log.scale = TRUE` returns all five PANSE parameters
(alpha, lambda, mean waiting time, NSE rate, NSE probability) for each codon:

```{r, echo = TRUE, eval = FALSE}
csp_nse <- getCSPEstimates(parameter_nse, samples = 500, log.scale = TRUE)
head(csp_nse$NSERate)
head(csp_nse$NSEProb)
```

### Partition function trace

The partition function Z normalises the per-gene expression model in PANSE.
Its trace can be inspected with:

```{r, echo = TRUE, eval = FALSE}
plot(trace_nse, what = "PartitionFunction")

pf_traces <- trace_nse$getPartitionFunctionTraces()  # list, one per mixture
pf_post   <- pf_traces[[1]][burnin_idx:length(pf_traces[[1]])]
cat("Z posterior mean:", mean(pf_post), "\n")
cat("Z 95% CI:        ", quantile(pf_post, c(0.025, 0.975)), "\n")
```

### Fixing parameters

To hold specific parameters constant (e.g., to test the PA model nested within
PANSE), use the locking methods:

```{r, echo = TRUE, eval = FALSE}
parameter_nse$fixAlpha()       # fix pausing shape
parameter_nse$fixLambdaPrime() # fix elongation rate
parameter_nse$fixNSERate()     # fix NSE rates (PANSE only)
parameter_nse$fixZ()           # fix partition function (PANSE only)
```

### Known limitations

- **Unpublished model**: PANSE has not undergone peer review. Results should
  be cross-validated against simulations before use in publications.
- **RFP count table size**: The model pre-computes a lgamma table for RFP
  counts 0-49 per codon. For highly expressed genes with counts >= 50, it falls
  back to runtime computation (correct, but slower).
- **Positional data required**: Unlike PA, PANSE needs `positional = TRUE`
  in `initializeGenomeObject` because cumulative survival probabilities depend
  on codon order within each gene.

## Restart files

For long runs it is useful to save periodic restart checkpoints so that
computation can be resumed after interruption:

```{r, echo = TRUE, eval = FALSE}
setRestartSettings(mcmc, file = "pa_restart.rst",
                   every = 50, initialize = TRUE)
runMCMC(mcmc, genome, model)
```

Pass the restart file to `initializeParameterObject` via the `restart.file`
argument to resume:

```{r, echo = TRUE, eval = FALSE}
parameter <- initializeParameterObject(model = "PA",
                                        restart.file = "pa_restart.rst")
```
