---
title: "AnaCoDa: Parameterizing the phi prior for ROC (anchoring, sphi, and HMC mixing)"
author: "Michael A. Gilchrist"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{AnaCoDa: Parameterizing the phi prior for ROC (anchoring, sphi, and HMC mixing)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

> **Scope.** This vignette covers the **single-lognormal** phi prior for the ROC
> model. It explains *why* the way you parameterize the prior on the gene-specific
> synthesis rate `phi` matters for sampling, and *how* to express your choice with
> the `phi.mphi` / `phi.sphi` arguments to `initializeParameterObject()`. The
> two-component (mixture / "dual-LN") prior is a different regime and is discussed
> only briefly at the end.

## The quantities you are specifying

ROC treats each gene's synthesis rate `phi_g` as lognormal,

$$\log \phi_g \sim \mathrm{Normal}(m_\phi,\; s_\phi),$$

so the prior has a **location** `mphi` ($m_\phi$, the mean of `log phi`) and a
**spread** `sphi` ($s_\phi$, the SD of `log phi`). The new API lets you specify
each independently:

```{r, eval = FALSE}
initializeParameterObject(
  genome,
  phi.mphi = constrained(statistic = "median", value = 1),   # the location
  phi.sphi = estimated(prior = prior_normal(mean = 1.4, sd = 0.125))  # the spread
)
```

The rest of this vignette is about choosing those two specs well. The short
version:

- **`phi.mphi`**: anchor a chosen *statistic* of `phi` to 1. Prefer
  `statistic = "median"` for fitting; use `"mean"` only for backward-compatibility /
  regression parity; **avoid `"mode"`**.
- **`phi.sphi`**: usually `estimated()` with a weakly-informative prior such as
  `prior_normal(mean = 1.4, sd = 0.125)`; `fixed()` when you want to hold the spread.
- For Stan/HMC fits, pair `statistic = "median"` with the **non-centered**
  parameterization and the **scale-anchor**, and keep the default **`diag_e`**
  metric (not `dense_e`).

## 1. Why parameterization matters: a non-identified scale

The ROC likelihood enters only through the product `dEta_c * phi_g` (in the linear
predictor `eta_{g,c} = -dM_c - dEta_c * phi_g`). Consequently it is **exactly
invariant** under the rescaling

$$\phi \to a\,\phi, \qquad dEta \to dEta / a \qquad (a > 0),$$

because `dEta * phi` is unchanged. The data therefore identify only the *product* of
the dEta scale and the phi scale -- **not the absolute level of `phi` on its own**.
Something other than the likelihood has to fix that level: that "something" is the
**anchor** (`phi.mphi`) plus the prior on `sphi`. In other words, the anchor is a
**gauge choice**, not a quantity the data estimate.

This non-identification is also the practical reason parameterization matters for
*sampling*: the rescaling direction is a near-flat ridge in the posterior, and it is
the dominant bottleneck for mixing the `dEta` (selection) parameters. The choices
below either fix the gauge in a numerically friendly way (Section 2-3) or reshape the
geometry so the sampler can move along/across that ridge efficiently (Section 4).

A useful corollary: because the level is a gauge, **`sphi` is gauge-invariant** and
the *relative* pattern of `dEta` across codons is gauge-invariant; only the *absolute*
`phi` level and the *absolute* `dEta` scale depend on the anchor. You can convert a
fit from one anchor to another after the fact (Section 2).

## 2. Anchoring the location: `phi.mphi = constrained(statistic, value = 1)`

`constrained(statistic, value = 1)` pins a chosen statistic of the `phi`
distribution to `value`. For the lognormal the three natural "central" statistics are

| statistic | value of the statistic | anchoring it to 1 forces |
|---|---|---|
| `"mean"`   | `exp(mphi + sphi^2/2)` | `mphi = -sphi^2/2` |
| `"median"` | `exp(mphi)`           | `mphi = 0` |
| `"mode"`   | `exp(mphi - sphi^2)`  | `mphi = +sphi^2` |

The key structural fact: **the choice of statistic is a choice of how `mphi` couples
to `sphi`.** `"median"` *decouples* them (`mphi = 0`, independent of `sphi`); `"mean"`
couples them weakly; `"mode"` couples them most strongly and in the opposite
direction.

What that does to the rest of the distribution (the API enforces these and warns
about the growth ones):

| anchor (=1) | E[phi] | median(phi) | mode(phi) |
|---|---|---|---|
| `mean`   | 1                    | `exp(-sphi^2/2)` | `exp(-3 sphi^2/2)` |
| `median` | `exp(sphi^2/2)`      | 1                | `exp(-sphi^2)` |
| `mode`   | `exp(3 sphi^2/2)`    | `exp(sphi^2)`    | 1 |

At a realistic `sphi = 1.4`, the contrast is stark -- where the *bulk* of `phi` mass
sits relative to 1:

| anchor (=1) | E[phi] | median(phi) |
|---|---:|---:|
| `mean`   | 1.00 | 0.38 |
| `median` | 2.66 | 1.00 |
| `mode`   | 18.9 | 7.1  |

### Recommendation

- **`"median"` (recommended for fitting).** Robust to the high-expression tail (the
  median ignores how extreme the most-expressed genes are), and it *decouples* `mphi`
  from `sphi`, which gives the cleanest sampling geometry and is what the Stan
  scale-anchor (Section 4) needs.
- **`"mean"` (legacy / regression parity).** The historical ROC convention
  (`E[phi] = 1`). Keep it when you need results numerically comparable to earlier ROC
  fits, or for regression test cases. It is sensitive to the tail (a few
  highly-expressed genes pull the level), and it couples `mphi` to `sphi`.
- **Avoid `"mode"`.** It couples `mphi` to `sphi` most strongly and in the direction
  that makes `E[phi]` *grow* as `exp(3 sphi^2/2)` -- for wide priors it pushes almost
  all of `phi` far above 1, fighting the `dEta` prior. `constrained("mode")` issues a
  warning to this effect.
- `"sd"` / `"variance"` pin the spread of `phi` rather than its center; they let
  `E[phi]` drift toward 0 as `sphi` grows and should be paired with a prior on `sphi`.
  The API warns accordingly.

```{r, eval = FALSE}
# Recommended for fitting:
phi.mphi <- constrained(statistic = "median", value = 1)

# Legacy / comparable-to-old-fits:
phi.mphi <- constrained(statistic = "mean", value = 1)

# Hold the location at an exact log-space mean instead of constraining a statistic:
phi.mphi <- fixed(value = 0)   # mphi = 0  (equivalent location to median=1)
```

### Re-gauging after a fit

Because the level is a gauge, a posterior fit under one anchor can be converted to
another deterministically. Going mean <-> median rescales `phi` and `dEta` by a
constant factor `exp(+/- sphi^2/2)`:

$$\phi_{\text{median}} = \phi_{\text{mean}}\,e^{+s_\phi^2/2}, \qquad
  dEta_{\text{median}} = dEta_{\text{mean}}\,e^{-s_\phi^2/2}.$$

So you can fit in the geometry-friendly `"median"` gauge and report in the legacy
`"mean"` gauge if you need comparability. `sphi` and the relative `dEta` pattern are
unchanged by the conversion.

## 3. The spread: `phi.sphi = estimated(prior = ...)` vs `fixed()`

```{r, eval = FALSE}
# Estimate sphi with a weakly-informative normal prior (recommended default):
phi.sphi <- estimated(prior = prior_normal(mean = 1.4, sd = 0.125))

# Estimate sphi with a bounded-uniform prior:
phi.sphi <- estimated(prior = prior_uniform(low = 0, high = 10))

# Estimate sphi with an improper flat prior (legacy default behavior):
phi.sphi <- estimated(prior = NULL)

# Hold sphi fixed:
phi.sphi <- fixed(value = 1.4)
```

`estimated()` puts `sphi` under MCMC with the given prior. The
`prior_normal(mean = 1.4, sd = 0.125)` form is the adopted weakly-informative default
for ROC fits -- centered near typical genome-wide `sphi` values and tight enough to
keep the sampler out of pathological small-`sphi` regions, but not so tight as to
override the data. `fixed()` is appropriate when you want to condition on a known
spread (e.g. transferring `sphi` from a reference fit, or a sensitivity analysis).

Currently implemented sphi priors are `prior_normal()`, `prior_uniform()`, and the
flat `NULL`. `prior_student_t()` and `prior_exponential()` are defined but **not yet
wired into estimation** (see Roadmap).

## 4. Sampling geometry for Stan/HMC fits

Fixing the gauge (Sections 2-3) makes the model identified, but a *residual* version
of the rescaling ridge still slows `dEta` mixing under HMC. Three settings address
the geometry; they are Stan data-flags on the ROC Stan model (`stan/roc_sphi_est.stan`)
and pair naturally with `phi.mphi = constrained("median")`.

**Non-centered `log_phi`.** Sample standardized `z` and set
`log_phi = mphi + sphi * z`. This removes the Neal's-funnel between `log_phi` and
`sphi` that otherwise appears once the global level is free, and it is *required* for
the scale-anchor below to actually decouple the ridge. (Trade-off: non-centering
slightly worsens `sphi` mixing; usually a good trade.)

**The dEta scale-anchor.** With a free median anchor, reparameterize so the global
level cancels out of the `dEta * phi` product. The likelihood then becomes exactly
invariant to the global level, which removes the dominant `dEta` ridge instead of
merely relocating it.

**Keep `diag_e`, not `dense_e`.** It is tempting to "let the metric learn the
correlations." Don't: the Stan mass matrix spans *all* sampled parameters -- roughly
`2 * (#codons) + (#genes) + 2`, on the order of 2000+ for a real genome -- and the
gene-by-gene block is essentially diagonal. A full `dense_e` therefore pays an
`O(n^2)` per-leapfrog cost dominated by modeling noise, with no benefit; the one
coupling that matters is the single low-rank rescaling direction, which the reparam
already removes. Use the default `diag_e`.

**Recommended HMC combination:**
`constrained("median")` + `estimated(sphi)` + non-centered + scale-anchor + `diag_e`.
On a `G = 2000` simulated yeast genome this lifted `dEta` effective-samples-per-second
to ~1.6x the centered/mean-anchored baseline, removed the funnel
(`log_phi` R-hat 2.49 -> ~1.0), and roughly halved warmup -- with no regression on
`dM` or `log_phi`. The reparameterization's correctness (exact invariance of the
likelihood to the global level) and its legacy-`mean` parity are covered by the
package tests `scripts/test_reparam_stan.R` and
`scripts/test_reparam_inertness_stan.R`.

## 5. Worked configurations

**(a) Legacy `mean(phi) = 1`, fixed spread** -- comparable to older ROC fits; good
for regression test cases.

```{r, eval = FALSE}
parameter <- initializeParameterObject(
  genome       = genome,
  model        = "ROC",
  num.mixtures = 1,
  gene.assignment = rep(1, length(genome)),
  phi.mphi = constrained(statistic = "mean", value = 1),
  phi.sphi = fixed(value = 1.4)
)
```

**(b) Recommended: `median(phi) = 1`, estimated spread** -- robust gauge, clean
geometry.

```{r, eval = FALSE}
parameter <- initializeParameterObject(
  genome       = genome,
  model        = "ROC",
  num.mixtures = 1,
  gene.assignment = rep(1, length(genome)),
  phi.mphi = constrained(statistic = "median", value = 1),
  phi.sphi = estimated(prior = prior_normal(mean = 1.4, sd = 0.125))
)
```

**Stan/HMC backend equivalents.** When fitting via the Stan ROC model rather than the
native MCMC, the same choices map to data-flags on `stan/roc_sphi_est.stan`:

```{r, eval = FALSE}
# anchor_phi        = 1     # median anchor (matches constrained("median"))
# noncentered       = 1     # non-centered log_phi
# deta_scale_anchor = 1     # cancel the global level from dEta*phi
# metric            = "diag_e"   # NOT dense_e
```

**Backward compatibility.** The older arguments still work and are translated
internally: `sphi = <number>` behaves like `phi.sphi = fixed(<number>)`,
`sphi = NA` like `phi.sphi = estimated()`, and
`phi.prior.constraint = "mean" | "median"` like
`phi.mphi = constrained(statistic = ...)`.

## 6. Caveats, scope, and roadmap

- **Single-LN only.** Everything above is for the one-component lognormal phi prior.
  The two-component mixture ("dual-LN") prior is a genuinely different regime: at
  realistic genome sizes the mixture is not reliably samplable by Stan HMC, and the
  native MCMC is the authoritative backend there. The anchoring ideas carry over only
  partially (the bulk component can be anchored, but the rare component and the
  mixing weight add their own identifiability issues).
- **The anchor is a convention.** Report and compare the *relative* `dEta` pattern
  (gauge-invariant). Absolute `phi` and absolute `dEta` magnitudes depend on the
  anchor; state which one you used.
- **Roadmap (not yet implemented).** `phi.mphi = estimated()` (estimating the global
  level under a prior rather than constraining it), and `prior_student_t()` /
  `prior_exponential()` for `sphi`, are defined in the API surface but currently stop
  with a "not yet implemented" message.
